{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-22 thesis/quicksand.tex
edit qs and tm stuff for consistency, clarity, typoity, etc

This chapter presents Quicksand,
an execution framework for model checking to manage this trade-off at run-time.
Given a fixed CPU budget,
representing the user's patience for testing,
Quicksand dynamically alters its preemption strategy based on data race analysis
(\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
and optimizes the size of state spaces on-the-fly,
guided by state space estimation
(\cite{estimation}, \cref{sec:landslide-estimate}),
to best match that budget.
I will discuss the trade-off inherent in number of preemption points used (\cref{sec:quicksand-pps}),
introduce {\em Iterative Deepening},
the algorithm that Quicksand uses to automatically navigate that trade-off
(\cref{sec:quicksand-id}),
prove its soundness relative to the more expensive full verification approach
(\cref{sec:quicksand-soundness}),
and present a large evaluation of Quicksand against several
state-of-the-art approaches % implemented under Landslide
in which Quicksand performs best on both bug-finding and verification
(\cref{sec:quicksand-eval}).
% justifying the claim that model checkers should navigate said tradeoff at runtime
}

@@

\qrevision{When configuring a model checker's preemption strategy,
or indeed,
choosing a model checker to begin with,
the resulting state space is {\em parameterized} by the set of preemption points.
%
In the first example of \Cref{fig:tree}
I constructed the state space by expanding both threads' {\tt x++;} operations into three pseudo-assembly instructions,
then designating every instruction as a possible preemption point,
yielding ${3+3 \choose 3} = 20$ total interleavings.
Later, \cref{sec:landslide-dpor} showed DPOR prune 16 of those as equivalent,
although the reduced state space size is still combinatorial
in the number of {\em conflicting} events, rather than the total number.
For larger tests, %with hundreds or thousands of conflicting events in any given execution,
committing in advance to test all possible interleavings quickly becomes impractical.
Accordingly, many existing model checkers opt for preempting on only a subset of execution events,
such as synchronization API boundaries.

@@

\begin{figure}[ht!] % TODO: check placement
	\begin{center}
		% FIXME: fix it to be a real tree
		\includegraphics[width=0.9\textwidth]{../oopsla/tree-maximal-only.pdf}
	\end{center}
	\caption[State space of \Cref{fig:pps-example} with synchronization preemption points only.]
	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
	Note that none of these interleavings preempt Thread 1 at the necessary place
	to trigger Thread 2's assertion failure..}
	\label{fig:pps-statespace}
\end{figure}
}

@@

Note that Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
Here
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

Iterative Deepening must be seeded with a set of initial state spaces,
which can be any number of subsets of the statically-available preemption points
that prior work model checkers would use.
The upcoming soundness proof relies on the maximal state space being included among these for verification's sake,
but to optimize for finding bugs faster,
implementations may wish to simultaneously to try testing subsets thereof.

For testing user-space code, Quicksand begins with the four jobs from \Cref{fig:id}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.

@@

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	\Input{$j_0$, the currently-running job}
	\Input{\AllJobs, the set of all existing (or completed) jobs}
	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
	\If{$\forall j \in \AllJobs,$
	\GetPPSet{$j_0$} $\cup$ $\alpha$
	$\not\subseteq$
	\GetPPSet{$j$}
	}{
		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
	}
	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
	}
	\caption{Adding new jobs with data-race preemption points.}
	\label{alg:handledatarace}
\end{algorithm}

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.8\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time,
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved access,
                and new jobs are added for later testing,
                corresponding to different combinations of those with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

I am presenting these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
written as sketches in informal prose,
to optimize for rapidly conveying an intuition for why it works
rather than to justify every internal step within the proof structure.%
\footnote{Also because I am already over 200 pages in this thesis, all told.}
% footnote yeah, i'm writing this last
% footnote of footnote yeah, the committee never read this part
A more rigorous treatment
is available in the tech report which accompanied the conference paper \cite{quicksand-soundness}.

@@

The available command-line options for configuring its
CPU-time budget, exploration modes, and so on
are listed in \cref{sec:landslide-quicksand-options}.
Additionally, Quicksand periodically issues a progress report
at fixed intervals to inform the user on the completion, bug-finding, and/or estimated progress of each job.
\Cref{fig:quicksand-progress} shows an example.
I highlight a few notable features of the jobs therein
to serve as a concrete example that may cement the reader's intuition
of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:
\begin{itemize}
	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
		which the user should examine to diagnose it.%
		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
		the other jobs come from a test with the bug already fixed,
		in order that exploration progress far enough to defer some jobs for the sake of example.
		In a real execution, jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
	\item Job 11 is the maximal state space, containing all synchronization preemption points
		and both currently-known data races.
	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
		at least twice as large as the provided CPU budget (1 hour),
		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
		This corresponds to the surprising ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
		for its ETA to be accurate enough yet.
\end{itemize}
Future work could extend these progress reports to be more interactive,
allowing the user to reprioritize state spaces at her whim
or to disable certain data-race preemption points after checking them by hand,
as discussed in \cref{sec:future-friendly}.

>>>>

This chapter presents Quicksand,
an execution framework for model checking to manage this trade-off at run-time.
Given a fixed CPU budget,
representing the user's patience for testing,
Quicksand dynamically alters its preemption strategy based on data race analysis
(\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
and optimizes the size of state spaces on the fly,
guided by state space estimation
(\cite{estimation}, \cref{sec:landslide-estimate}),
to best match that budget.
I will discuss the trade-off inherent in number of preemption points used (\cref{sec:quicksand-pps}),
introduce {\em Iterative Deepening},
the algorithm that Quicksand uses to automatically navigate that trade-off
(\cref{sec:quicksand-id}),
prove its soundness relative to the more expensive full verification approach
(\cref{sec:quicksand-soundness}),
and present a large evaluation of Quicksand against several
state-of-the-art approaches % implemented under Landslide
in which Quicksand performs best on both bug-finding and verification
(\cref{sec:quicksand-eval}).
% justifying the claim that model checkers should navigate said tradeoff at runtime
}

@@

\qrevision{When configuring a model checker's preemption strategy,
or indeed,
choosing a model checker to begin with,
the resulting state space is {\em parameterized} by the set of preemption points.
%
In the first example of \Cref{fig:tree}
I constructed the state space by expanding both threads' {\tt x++} operations into three pseudo-assembly instructions,
then designating every instruction as a possible preemption point,
yielding ${3+3 \choose 3} = 20$ total interleavings.
Later, \cref{sec:landslide-dpor} showed DPOR prune 16 of those as equivalent,
although the reduced state space size is still combinatorial
in the number of {\em conflicting} events, rather than the total number.
For larger tests, %with hundreds or thousands of conflicting events in any given execution,
committing in advance to test all possible interleavings quickly becomes impractical.
Accordingly, many existing model checkers opt for preempting on only a subset of execution events,
such as synchronization API boundaries.

@@

\begin{figure}[ht!] % TODO: check placement
	\begin{center}
		% FIXME: fix it to be a real tree
		\includegraphics[width=0.9\textwidth]{../oopsla/tree-maximal-only.pdf}
	\end{center}
	\caption[State space of \Cref{fig:pps-example} with synchronization preemption points only.]
	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
	Note that none of these interleavings preempt Thread 1 at the necessary place
	to trigger Thread 2's assertion failure.}
	\label{fig:pps-statespace}
\end{figure}
}

@@

%Note that
Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
This chapter
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

% TODO: fix this section, the way it lists the pps is fucked, figure reference prob dnagling
Iterative Deepening must be seeded with a set of initial state spaces,
which can be any number of subsets of the statically-available preemption points
that prior work model checkers would use.
The upcoming soundness proof relies on the maximal state space being included among these for verification's sake,
but to optimize for finding bugs faster,
implementations may wish to simultaneously to try testing subsets thereof.

For testing user-space code, Quicksand begins with the four
% TODO: decide
jobs from \Cref{fig:id}:
%possible combinations of preemption points from \Cref{fig:pps-statespace}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.

@@

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[h]
	\SetKwInOut{Input}{Input}
	\Input{$j_0$, the currently-running job}
	\Input{\AllJobs, the set of all existing (or completed) jobs}
	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
	\If{$\forall j \in \AllJobs,$
	\GetPPSet{$j_0$} $\cup$ $\alpha$
	$\not\subseteq$
	\GetPPSet{$j$}
	}{
		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
	}
	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
	}
	\caption{Adding new jobs with data-race preemption points.}
	\label{alg:handledatarace}
\end{algorithm}

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved memory access,
                and new jobs are added for later testing,
                corresponding to different combinations thereof with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

I present these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
written as sketches in informal prose,
to optimize for rapidly conveying an intuition for why it works
rather than to justify every internal step within the proof structure.%
\footnote{Also because I am already over 200 pages in this thesis, all told.}
% footnote yeah, i'm writing this last
% footnote of footnote yeah, the committee never read this part
A more rigorous treatment
is available in the tech report which accompanied the conference paper \cite{quicksand-soundness}.

@@

The available command-line options for configuring its
CPU-time budget, exploration modes, and so on
are listed in \cref{sec:landslide-quicksand-options}.
Additionally, Quicksand periodically issues a progress report
at fixed intervals to inform the user on the completion, bug-finding, and/or estimated progress of each job.
\Cref{fig:quicksand-progress} shows an example.
I highlight a few notable features of the jobs therein
to serve as a concrete example that may cement the reader's intuition
of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:
\begin{itemize}
	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
		which the user should examine to diagnose it.%
		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
		the other jobs come from a test with the bug already fixed,
		in order that exploration progress far enough to defer some jobs for the sake of example.
		In a real execution, the superset jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
	\item Job 11 is the maximal state space, containing all synchronization preemption points
		and both currently-known data races.
	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
		at least twice as large as the provided CPU budget (1 hour),
		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
		This corresponds to the surprising ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
		for its ETA to be accurate enough yet.
\end{itemize}
Future work could extend these progress reports to be more interactive,
allowing the user to reprioritize state spaces at her whim
or to disable certain data-race preemption points after checking them by hand,
as discussed in \cref{sec:future-friendly}.

### 2018-12-22 thesis/tm.tex
few tweaks across chapters, use less subsubsection
edit qs and tm stuff for consistency, clarity, typoity, etc

\subsubsection{Experimental setup}
\label{sec:tm-eval-exp-setup}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{l}
		\texttt{\flow{while} (\_retry);} \\
		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\call{\_xend}();} \\
		\texttt{\} \flow{else} \{} \\
		\texttt{~~~~\call{pthread\_mutex\_lock}(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = \const{true};} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\_retry = \const{false};} \\
		\texttt{~~~~\call{pthread\_mutex\_unlock}(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption[Code from {\tt htmavl.hpp} showing the segfault bug Landslide found.]
		{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown segfault bug
		Landslide found in {\tt avl\_insert}.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}

\begin{itemize}
	\item
	{\bf Atomicity violation.}
	{\tt avl\_insert} with any parameters higher than (2,1)
	exposed a previously-unknown bug in the transactional AVL tree.
	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
	This manifested
	alternately
	% landslide-trace-1533850116.527842.html
	% landslide-trace-1533850236.873032.html
	as a segfault (for test parameters (2,2) and (3,1))
	and
	% landslide-trace-1533854371.201589.html
	as a consistency-check assertion failure (for test parameters (2,3)).
	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
	(between it and {\tt \_xbegin()}),
	making the bug extremely unlikely to manifest under stress testing,
	but Landslide is blind to such matters of chance.

@@

\subsubsection{Retry set optimality}

@@

\subsubsection{STM reduction soundness}

@@

\subsubsection{Nested transactions}
\label{sec:tm-warpzone-nested}

@@

\subsubsection{Relaxed memory orderings}
\label{sec:tm-warpzone-relaxed}

>>>>

\subsection{Experimental setup}
\label{sec:tm-eval-exp-setup}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{l}
		\texttt{\flow{while} (\_retry) \flow{continue};} \\ % in the actual code it's "while (retry);"; i'm adding continue for clarity
		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\call{\_xend}();} \\
		\texttt{\} \flow{else} \{} \\
		\texttt{~~~~\call{pthread\_mutex\_lock}(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = \const{true};} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\_retry = \const{false};} \\
		\texttt{~~~~\call{pthread\_mutex\_unlock}(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption[Code from {\tt htmavl.hpp} showing the segfault bug Landslide found.]
		{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown segfault bug
		Landslide found in {\tt avl\_insert}.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}

\begin{itemize}
	\item
	{\bf Atomicity violation.}
	{\tt avl\_insert} with any parameters higher than (2,1)
	exposed a previously-unknown bug in the transactional AVL tree.
	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
	This manifested
	alternately
	% landslide-trace-1533850116.527842.html
	% landslide-trace-1533850236.873032.html
	as a segfault (for test parameters (2,2) and (3,1))
	and
	% landslide-trace-1533854371.201589.html
	as a consistency-check assertion failure (for test parameters (2,3)).
	The presence of {\tt while (\_retry) continue} makes the necessary preemption window extremely small
	(between it and {\tt \_xbegin()}),
	making the bug extremely unlikely to manifest under stress testing,
	but Landslide is blind to such matters of chance.

@@

\subsection{Retry set optimality}

@@

\subsection{STM reduction soundness}

@@

\subsection{Nested transactions}
\label{sec:tm-warpzone-nested}

@@

\subsection{Relaxed memory orderings}
\label{sec:tm-warpzone-relaxed}
